---
title: Dataframes
teaching: 25
exercises: 15
questions:
  - "What is a data frame?"
  - "Why use a data frame as a tidy data structure?"
objectives:
  - "To learn how to create a data frame."
  - "To understand how to find basic information about a data frame"
  - "To know how to inspect the data in a data frame"
keypoints:
  - "Data frames (or tibbles in the tidyverse) are lists where each element is a vector of the same length"
  - "Create them with `data_frame()`"
  - "Use `nrow()`, `ncol()`, `dim()`, or `colnames()` to find information about a data frame"
  - "Use `head()`, `tail()`, `summary()`, or `glimpse()` to inspect a data frame's content"
source: Rmd
---

```{r, include=FALSE}
source("../bin/chunk-options.R")
knitr_fig_path("06-")

library(tidyverse)
options(max.print = 80)
```

## Dataframes
List of vectors

## Dataframes as a Tidy structure

## Structor of the dataframe

## Inspecting the data
* Head/Tail
* Glimpse
* Summary
* Class

---

Now that we understand a little bit about why we might prefer our data to be 'tidy', we have one data
structure left to learn - the data frame. Data frames are a special type of list comprised of vectors
that all have to be the same length. This means that a data frame produces a rectangular table of data,
as shown below using the `diamonds` data set that is loaded as part of the tidyverse. Data frames are
an ideal format for storing and working with tidy data and all the tidyverse tools we will be learning
from here are designed to work with data in this form.

```{r tibble}
diamonds
```

> ## What's a tibble?
> You might notice in the output above that it calls itself a tibble, rather than a data frame.
> A tibble is just the tidyverse's version of a data frame that has a few behaviours tweaked to 
> make it behave more predictibly. Try comparing the output of a base R `data.frame` version of 
> diamonds with `as.data.frame(diamonds)` to get some idea of the differences.
> 
> We will try to refer to them as data frames throughout these lessons. But know that data frames
> and tibbles are interchangable for our purposes.
{: .callout}


## Inspecting a data frame

Looking at the printed output from the `diamonds` data frame can tell us a lot of information about
it. The first line tells us the dimensions of the data, in this case there are 53,940 rows and 10 
columns. This information can also be found with:
```{r data-frame-dimensions}
nrow(diamonds)
ncol(diamonds)
dim(diamonds)
```

The next row of output gives the names of the columns, which can also be found using `colnames()`.

```{rdata-frame-colnames}
colnames(diamonds)
```

The next row tells you the data type of each column, followed by the data itself.

Because data frames are just lists underneath, you can use the same `$` or `[[` notation to access just the
data from a single column:

```{r df-dollar}
# Use $ to access by name
diamonds$price

# Use [[ ]] to access by name or position 
diamonds[["price"]]
diamonds[[7]]
```

While it might seem confusing, data frames can also be subset using `[`. In this case, you call it 
using `data_frame[row_selector , column_selector]`. Where `row_selector` and `column_selector` can
be a set of numeric indexes or names to extract, or a vector of logical TRUE/FALSEs that show which
rows/column to keep.

```{r df-subset}
#First 5 rows
diamonds[1:5, ]

#First 5 columns, don't forget the comma
diamonds[, 1:5]

#Rows 1, 3 and 5 with the carat and price columns
diamonds[c(1, 3, 5), c("carat", "price")]
```

## Overview of a data frame
There are many other ways to view the data and look at its data types, and the
structure of the data. To look at the first 5 rows of the diamonds dataset, use `head()`:

```{r df-head}
head(diamonds, 5)
```

and use `tail()` to look at the last 10 rows:

```{r df-tail}
tail(diamonds, 10)
```

To look at the structure of the data (particularly when there are many columns) use `glimpse()`:

```{r df-glimpse}
glimpse(diamonds)
```

And use `summary()` to get a summarised breakdown of each column:

```{r df-summary}
summary(diamonds)
```

> ## Challenge 1
> What does the output from `summary(diamonds)` show you? Why do you think it has a different format
> for different columns?
> > ## Solution to Challenge 1
> > The output from `summary()` changes depending on the class of the data in the column. For the numeric
> > columns it shows the minimum, maximum, mean and quartile values. For the others, it shows a count 
> > of the contents.

## Factors

The `cut`, `color`, and `clarity` columns in the `diamonds` data set might be a little confusing. What
is this `<ord>` type, and why are they summarised as a count of the individual terms? These columns 
are each *factors* (specifically ordered factors, hence the `<ord>`), which usually look like character 
data, but are typically used to represent categorical information.

```{r}
class(diamonds$cut)
diamonds$cut
```

Here you can see that the `cut` column in the `diamonds` data set appears to be a character vector,
listing the quality of the diamond's cut. In the final line, you can see that there are a set of 
levels of this factor: "Fair < Good < Very Good < Premium < Ideal". The levels of this factor can
also be accessed using:

```{r df-levels}
levels(diamonds$cut)
```
 
But what happens when we look more closely at this data using `glimpse()`

```{r glimpse-factor}
glimpse(diamonds$cut)
```

This tells us we have an ordered factor with 5 levels, just like we expected. But when it comes to 
show the data itself, all we see are a bunch of numbers. This is because, to R, a factor is really 
just an integer underneath, with the levels telling it how to map the integer to the actual category.
So a value of 1 would map to the first level (`Fair`), while a value of 4 would map to the fourth
level (`Premium`).

Expecting factors to behave as characters, rather than integers, is a common cause of errors for 
people new to R. So always remember to inspect your data with the methods shown here to make sure it
is of the right type.

## Your turn
So far, you've been walked through investigating a data frame. Let's use those skills to explore a
data set you have not yet been exposed to. 

> ## Challenge 2
>
> The `storms` data set contains information on hurricanes recorded in the Atlantic Ocean. Using the 
> tools you have learned so far, explore this data set and describe what it contains. Explain both
> the structural features of the data set as a whole, as well as it's content.
>
> > ## Solution to Challenge 2
> >
> > The object `storms` is a data frame (a tibble) with 10,010 rows and 13 columns.
> > - `name` and `status` are character vectors.
> > - `year`, `month`, `hour`, `lat`, `long`, `ts_diamater` and `hu_diameter` are numeric vectors.
> > - `day`, `wind`, and `pressure` are integer vectors.
> > - `category` is an ordered factor vector, with levels `-1 < 0 < 1 < 2 < 3 < 4 < 5`
> >
> {: .solution}
{: .challenge}
