---
title: Extract rows
teaching: 15
exercises: 10
questions:
  - "How can I extract rows from a data frame based on their values?"
objectives:
  - "Filter data frames using logical operators"
  - "Combine logical expressions in a filter"
keypoints:
  - "Use `filter()` to choose data based on values."
  - "The logical operator `%in%` can filter data from a list of possible values"
source: Rmd
---

```{r, include=FALSE}
source("../bin/chunk-options.R")
knitr_fig_path("07-")

library(tidyverse)
gapminder <- read_csv("data/gapminder.csv")
```

If we are looking to subset the rows of our data (rather than the columns using `select()`) we can 
use `filter()`. This function takes a data frame as it's first argument, then a set of conditions to
check. Any row of the data frame that meets all the conditions is kept and any that fail are discarded.

For example, to get just the Australian data from the gapminder set:

```{r filter-intro}
filter(gapminder, country == "Australia")
```

or to get only data from 1997 or later:

```{r filter-gt}
filter(gapminder, year >= 1997)
```

## Filter operators
Any of the standard (comparison operators)({{ page.root }}{% link _episodes/02-Using-R.md %}#comparing-things)
can be used in a filter

  * `==` equal to
  * `!=` not equal to
  * `<`/`>` less/greater than
  * `<=`/`>=` less/greater than or equal to

Any function that produces a `TRUE`/`FALSE` output (such as the `is.na()` function) can be used as well.

> ## Challenge 1
> Extract all rows from `gapminder` where the recorded life expectancy is 80 or higher.
>
> Then try to extract just the rows from European countries.
> > ## Solution to Challenge 1
> > ``` {r filter-sol-1}
> > filter(gapminder, lifeExp >= 80)
> >
> > filter(gapminder, continent == "Europe")
> > ```
> {: .solution}
{: .challenge}

If you provide multiple filter conditions (separated by a comma) then only rows matching **all** of
the conditions will be kept.

```{r filter-combine-conditions}
filter(gapminder, country == "Australia", year >= 1997)
```

> ## Challenge 2
> How would you extract just rows that are healthy (life expectancy of 80 or higher) **and** rich
> (GDP per capita of $30,000 or higher)?
> > ## Solution to Challenge 2
> > ``` {r filter-sol-2}
> > filter(gapminder, lifeExp >= 80, gdpPercap >= 30000)
> > ```
> {: .solution}
{: .challenge}

If you need to keep all rows that meet one or the other condition, use the logical OR operator (`|`).
`|` resolves as `TRUE` if either of the left and right conditions are `TRUE`.

```{r filter-or}
# Tries to find data from Australia AND New Zealand (and returns an empty table)
filter(gapminder, country == "Australia", country == "New Zealand")

# Use | to look for data from Australia OR New Zealand
filter(gapminder, country == "Australia" | country == "New Zealand")
```

It might seem natural to write `country == "Australia" | "New Zealand"` but try it and you will see
that you get an error. Each side of the `|` operator must result in a `TRUE` or `FALSE`, and "New
Zealand" is neither. 

> ## Challenge 3
> Modify your answer to Challenge 2 so that you extract the rows where countries are **either**
> healthy (life expectancy of 80 or higher) or rich (GDP per capita of $30,000 or higher).
> > ## Solution to Challenge 3
> > ``` {r filter-sol-3}
> > filter(gapminder, lifeExp >= 80 | gdpPercap >= 30000)
> > ```
> {: .solution}
{: .challenge}

When you have many possible matches from a row that you want to keep, writing a long expression with 
many `|` can be time consuming and error prone. Instead, the `%in%` operator can be used to simplify
things. `%in%` returns `TRUE` if the left hand side is found somewhere in the right hand side and
`FALSE` otherwise.

```{r filter-in}
filter(gapminder, country %in% c("Australia", "New Zealand"))
```

> ## Challenge 4
> Extract the rows from all countries in Africa, Asia, or Europe. How many rows does your dataframe 
> have?
> > ## Solution to Challenge 4
> > ``` {r filter-sol-4}
> > filter(gapminder, continent %in% c("Africa","Asia", "Europe"))
> > ```
> {: .solution}
{: .challenge}


