---
title: Selecting columns
teaching: 15
exercises: 10
questions:
  - "How can I select specific columns from a data frame to work with?"
objectives:
  - "Select columns using several methods"
  - "Rename columns to make them easier to work with"
keypoints:
  - "Use `select()` to choose variables from a dataframe."
  - "Helper functions make it easier to select the correct columns."
  - "Use `rename()` to rename variables without dropping columns."
source: Rmd
---

```{r, include=FALSE}
source("../bin/chunk-options.R")
knitr_fig_path("08-")

library(tidyverse)
```

The first task we will cover is extracting specific columns from a dataset to work with. This 
may be needed if your data is very, very wide and you are only interested in a few columns. The 
gapminder data doesn't exactly meet that description, but we will continue using it to understand 
how the functions work.

```{r load-gapminder, message=F}
gapminder <- read_csv("data/gapminder.csv")
```

## `select()`

You can specify columns to keep from a data frame using `select()`. At it's simplest, you provide 
`select()` with a data frame and the column names you would like to keep.

```{r select-by-name}
#Select the year, country and pop columns
select(gapminder, year, country, pop)
```

Here you can see that the `select()` produces a new data frame with just the columns you provided. 
It also keeps them in the order you specified.

You can also select columns by index position:

```{r select-by-pos}
#Select the first, third and fourth columns
select(gapminder, 1, 3, 4)
```

or you can select columns to *exclude* by using negation:

```{r select-by-negation}
#Remove lifeExp and pop columns
select(gapminder, -lifeExp, -pop)
```

If the columns you want are all together in the original data frame, you can select them as a range

```{r select-range}
#Select all columns between year and lifeExp
select(gapminder, year:lifeExp)

#Exclude all columns between year and lifeExp
select(gapminder, -(year:lifeExp))
```

> ## Save your output
> In the examples above, the output of the `select()` function is being printed to the screen to 
> demonstrate how it works. Normally, you will want to actually *use* the selected data later on, so
> don't forget to save the output as a variable.
> ```{r select-save-variable}
> # Just prints to screen, can't use data
> select(gapminder, year, country, pop)
> 
> # Saves output to variable; can be accessed later
> just_population <- select(gapminder, year, country, pop)
> 
> just_population
> ```
{: .callout}


> ## Challenge 1
> Use three different methods to select just the `country`, `year`, `pop`, and `gdpPercap` columns 
> from the `gapminder` dataframe.
> > ## Solution to Challenge 1
> > ```{r select-sol-1}
> > #By name
> > select(gapminder, country, year, pop, gdpPercap)
> >
> > #By position
> > select(gapminder, 1, 3, 5, 6)
> > 
> > #By exclusion
> > select(gapminder, -continent, -lifeExp)
> > ```
> {: .solution}
{: .challenge}

## Select helper functions

There are a number of helper functions that can be used to select the correct columns. Some commonly
used ones include `starts_with()`, `ends_with()` and `contains()`, but you can see a full list by 
looking at the help file (`?tidyselect::select_helpers`).

These helpers do exactly what you would expect from their names, but to see them in action:

```{r select-helpers}
#All columns that start with "co"
select(gapminder, starts_with("co"))

#All columns that contain the letter "e"
select(gapminder, contains("e"))
```

> ## Challenge 2
> Select all columns from `gapminder` that end with the letter "p"
>
> > ## Solution to Challenge 2
> > ```{r select-sol-2}
> > select(gapminder, ends_with("p"))
> > ```
> {: .solution}
{: .challenge}

> ## Challenge 3
> Run the following code:
> ~~~~
> select(gapminder, contains("P"))
> ~~~~
> {: .language-r}
>
> Does it behave the way you expected? What additional argument do you need to provide to `contains()`
> to make it return only the `gdpPercap` column?
> > ## Solution to Challenge 3
> > ```{r select-sol-3}
> > select(gapminder, contains("P"))
> >
> > select(gapminder, contains("P", ignore.case = F))
> > ```
> {: .solution}
{: .challenge}

## `rename()`

`select()` *can* be used to rename columns while you are selecting them using the form 
`new_name = old_name`:

```{r select-rename}
select(gapminder, country_name = country, population = pop)
```

but since it drops all the other columns, it is might not be the outcome you were looking for. 
Instead, `rename()` can rename columns while retaining the rest of the data frame.

```{r rename-example}
rename(gapminder, country_name = country, population = pop)
```

This will often be useful to convert column names from your imported data into the 
[consistent naming format]({{ page.root }}{% link _episodes/02-Using-R.md %}#naming-things) you
have decided on.

> ## Challenge 4
> Rename the `lifeExp` and `gdpPercap` columns to the `snake_case` format
>
> > ## Solution to Challenge 4
> > ```{r select-sol-4}
> > rename(gapminder, life_exp = lifeExp, gdp_per_cap = gdpPercap)
> > ```
> {: .solution}
{: .challenge}