---
title: Data Verbs - select 
teaching: 
exercises: 
questions:
objectives:
keypoints:
source: Rmd
---

```{r, include=FALSE}
source("../bin/chunk-options.R")
knitr_fig_path("09-")

library(tidyverse)
```

The first task we will cover is extracting specific columns from a dataset to work with. This 
may be needed if your data is very, very wide and you are only interested in a few columns. The 
gapminder doesn't exactly meet that description, but we will continue using it to understand how the
functions work.

```{r load-gapminder, message=F}
gapminder <- read_csv("data/gapminder.csv")
```

## `select()`

You can specify columns to keep from a data frame using `select()`. At it's simplest, you provide 
`select()` with a data frame and the column names you would like to keep.

```{r select-1}
select(gapminder, country, year, pop)
```





  * to select specific columns from a dataframe 
    * by name
    * by position
    * by negation
    * functions (eg. `everything()`?)

## `rename` 
  * to rename columns
  * link back in to discussion of naming requirements/conventions in episode 2
  


