---
title: Selecting rows
teaching: 15
exercises: 10
questions:
  - "How can I select rows from a data frame based on their values?"
objectives:
  - "Filter data frames using logical operators"
  - "Combine logical expressions in a filter"
keypoints:
  - "Use `filter()` to choose data based on values."
  - "The logical operator `%in%` can filter data from a list of possible values"
source: Rmd
---

```{r, include=FALSE}
source("../bin/chunk-options.R")
knitr_fig_path("10-")

library(tidyverse)
gapminder <- read_csv("data/gapminder.csv")
```

If we are looking to subset the rows of our data (rather than the columns using `select()`) we can 
use `filter()`. This function takes a data frame as it's first argument, then a set of conditions to
check. Any row of the data frame that meets all the conditions is kept and any that fail are discarded.

For example, to get just the Australian data from the gapminder set:

```{r filter-intro}
filter(gapminder, country == "Australia")
```

or to get only data from 1997 or later:

```{r filter-gt}
filter(gapminder, year >= 1997)
```

## Filter operators
Any of the standard (comparison operators)({{ page.root }}{% link _episodes/02-Using-R.md %}#comparing-things)
can be used in a filter

  * `==` equal to
  * `!=` not equal to
  * `<`/`>` less/greater than
  * `<=`/`>=` less/greater than or equal to

If you provide multiple filter conditions (separated by a comma) then only rows matching **all** of
the conditions will be kept.

> ## Challenge 1
> How would you extract just the Australian data from 1997 or later?
> > ## Solution to Challenge 1
> > ``` {r filter-sol-1}
> > filter(gapminder, country == "Australia", year >= 1997)
> > ```
> {: .solution}
{: .challenge}

If you need to keep all rows that meet one or the other condition, use the logical OR operator (`|`)

```{r filter-or}
# Tries to find data from Australia AND New Zealand (and returns an empty table)
filter(gapminder, country == "Australia", country == "New Zealand")

# Use | to look for data from Australia OR New Zealand
filter(gapminder, country == "Australia" | country == "New Zealand")
```

> ## Challenge 2
> Modify your answer to Challenge 1 so that you extract all data from Australia, or from 1997 and 
> later.
> > ## Solution to Challenge 2
> > ``` {r filter-sol-2}
> > filter(gapminder, country == "Australia" | year >= 1997)
> > ```
> {: .solution}
{: .challenge}

It might seem natural to write `country == "Australia" | "New Zealand"` but try it and you will see
that you get an error. Each side of the `|` operator must result in a `TRUE` or `FALSE`, and "New
Zealand" is neither.

When you have many possible matches from a row that you want to keep, writing a long expression with 
many `|` can be time consuming and error prone. Instead, the `%in%` operator can be used to simplify
things. `%in%` returns `TRUE` if the left hand side is found somewhere in the right hand side and
`FALSE` otherwise.

```{r filter-in}
filter(gapminder, country %in% c("Australia", "New Zealand"))
```

> ## Challenge 3
> Extract the rows from all countries in Africa, Asia, or Europe. How many rows does your dataframe 
> have and why?
> > ## Solution to Challenge 3
> > ``` {r filter-sol-3}
> > filter(gapminder, continent %in% c("Africa","Asia", "Europe"))
> > ```
> {: .solution}
{: .challenge}


