---
title: Summarise and Grouping
teaching: 20
exercises: 20
questions:
  - "How can I make summaries of groups of data?"
  - "How can I count the elements of groups in a dataset?"
objectives:
  - "Identify when grouping is necessary to make summaries"
  - "Be able to to usefully summarise variables in a dataframe"
keypoints:
  - "`summarise()` creates a new variable that provides a one row summary per group"
  - "`n()` is useful to count rows per group"
source: Rmd
---

```{r, include=FALSE}
source("../bin/chunk-options.R")
knitr_fig_path("09-")

library(tidyverse)
gapminder <- read_csv("data/gapminder.csv")
```

The `summarise()` function lets us create new variables by collapsing a data frame into a single 
summary statistic. You can use `summarise()` with any function that takes a vector as input and 
returns a single value as output. For example, what is the average life expectance in the gapminder 
data?

```{r summarise-intro}
summarise(gapminder, mean_life_exp = mean(lifeExp))
```

On it's own, this may not seem that exciting. You could just as easily get the same result by using
`mean(gapminder$lifeExp)`. Where is becomes more useful however, is that you can use multiple 
summary functions at the same time.

```{r summarise-many-funcs}
summarise(
    gapminder,
    mean_life_exp = mean(lifeExp),
    sd_life_exp = sd(lifeExp),
    mean_gdp_per_cap = mean(gdpPercap),
    max_gdp_per_cap = max(gdpPercap)
)
```

> ## Challenge 1
> Calculate the mean and median population for the gapminder data
> > ## Solution to Challenge 1
> > ``` {r summarise-sol-1}
> > summarise(gapminder, mean_pop = mean(pop), median_pop = median(pop))
> > ```
> {: .solution}
{: .challenge}

and you can get summaries for different groups in conjunction with `group_by()`

```{r summarise-groups}
gapminder_by_country <- group_by(gapminder, country)

summarise(gapminder_by_country, mean_life_exp = mean(lifeExp))
```

> ## Challenge 2
> Adjust your answer to Challenge 1 to show the mean and median population for each continent.
> > ## Solution to Challenge 2
> > ``` {r summarise-sol-2}
> > gapminder_by_continent <- group_by(gapminder, continent)
> > 
> > summarise(gapminder_by_continent, mean_pop = mean(pop), median_pop = median(pop))
> > ```
> {: .solution}
{: .challenge}

## Sorting your results
If you need to sort your resulting data frame by a particular variable, use `arrange()`. This 
function takes a data frame and a set of column names and it rearranges the rows so that the 
specified columns are in order.

```{r summarise-arrange}
arrange(gapminder, gdpPercap)

# Use desc() to sort from highest to lowest
arrange(gapminder, desc(gdpPercap))
```

> ## Challenge 3
> Calculate the average life expectancy per country. Which has the shortest average life expectancy 
> and which has the longest average life expectancy?
> > ## Solution to Challenge 3
> > ``` {r summarise-sol-3}
> > summarised_life_exp <- summarise(gapminder_by_country, mean_life_exp = mean(lifeExp))
> >
> > arrange(summarised_life_exp, mean_life_exp)
> >
> > arrange(summarised_life_exp, desc(mean_life_exp))
> > ```
> {: .solution}
{: .challenge}

## Counting things
A very common summary operation is to count the number of observations. The `n()`
function will help simplify this process. `n()` will return the number of rows in the data frame (or
in the group if the data frame is grouped).

```{r using-n}
summarise(gapminder, num_rows = n())

summarise(gapminder_by_country, num_rows = n())
```

The `n()` function can be very useful if we need to use the number of observations in calculations.
For instance, if we wanted to get the standard error of the population per country:

```{r counting-n}
#standard error = standard deviation / square root of the number of samples
summarise(gapminder_by_country, se_pop = sd(pop) / sqrt(n()) )
```

> ## Challenge 4
> Let's try to put together all three functions introduced here. Produce a data frame that summarises
> the number of rows for each continent, sorted from highest to lowest. Use 
> `group_by()`, `summarise()`, and `arrange()` in that order to achieve it.
> > ## Solution to Challenge 4
> > ``` {r summarise-sol-4}
> > gap_by_cont <- group_by(gapminder, continent) 
> > 
> > count_by_cont <- summarise(gap_by_cont, num_rows = n())
> > 
> > arrange(count_by_cont, desc(num_rows))
> > ```
> {: .solution}
{: .challenge}
