---
title: Creating New Variables
teaching: 30
exercises: 15
questions:
  - "How can I create new variables in my data frame?"
  - "How do I deal with a data frame made up of different groups?"
objectives:
  - "Create new columns by performing calculations on old variables."
  - "Use functions to create new variables."
keypoints:
  - "Use `mutate()` to create new variables from old ones."
  - "You can create new variables using any function that returns a vector of the same length as the data frame."
  - "Use `group_by()` to group your data based on a variable."
source: Rmd
---

```{r, include=FALSE}
source("../bin/chunk-options.R")
knitr_fig_path("09-")

library(tidyverse)
gapminder <- read_csv("data/gapminder.csv")
```

So far, we have been looking at functions that deal just with the variables in our data frame. But
what if we are needing to create *new* variables? For that task, we will use the `mutate()` function
from the [dplyr](https://dplyr.tidyverse.org) package of the tidyverse.

To create a new variable, you provide `mutate()` with the name of the new column and how to calculate
the new value.

```{r mutate-intro}
mutate(gapminder, gdp = gdpPercap * pop)
```

This line adds a new column to our gapminder data called `gdp` and the value in this column is 
calculated by multiplying the `gdpPercap` and the `pop` figure for each row.

> ## Challenge 1
> Create a new column in the `gapminder` data frame that contains the population in millions (ie.
> divide the population column by 1,000,000).
> > ## Solution to Challenge 1
> > ```{r mutate-sol-1}
> > mutate(gapminder, pop_m = pop/1e6)
> > ```
> {: .solution}
{: .challenge}

You are not limited to mathematical operators in creating new columns. Any function that produces a
value for each row can be used:

```{r mutate-function}
# Take the logarithm of the population value
mutate(gapminder, log_pop = log(pop))

# Abbreviate the country name
# str_sub() takes a subset of a string from the given coordinates
mutate(gapminder, country_abbr = str_sub(country, start = 1, end = 4))
```

> ## Challenge 2
> Using the function `str_length()` which returns the length of a string, create a new column that 
> contains the number of letters in the country's name.
> > ## Solution to Challenge 2
> > ```{r mutate-sol-2}
> > mutate(gapminder, num_letters = str_length(country))
> > ```
> {: .solution}
{: .challenge}

In fact, anything that produces a vector of the same length as the data frame can be used. It does 
not even have to reference data within the data frame.

```{r mutate-new-column}
# 1704 rows in gapminder
index_numbers <- 1:1704

mutate(gapminder, row_num = index_numbers)
```

## Multiple mutations
You can perform multiple mutations at the same time by separating them with a comma.

```{r mutate-multiple}
mutate(gapminder, gdp = gdpPercap * pop, log_pop = log(pop))
```

One useful feature is that you can refer to your created columns later on in the same
`mutate()` call.

```{r mutate-chaining}
mutate(
  gapminder,
  gdp = gdpPercap * pop,
  log_gdp = log(gdp)
)
```

> ## Challenge 3
> Create a column in the gapminder data showing the life expectancy in days and one for GDP in billions
> of dollars
> > ## Solution to Challenge 3
> > ``` {r mutate-sol-3}
> > gapminder %>% 
> >   mutate(
> >     life_exp_days = lifeExp * 365,
> >     gdp_billion = gdpPercap * pop / 10^9
> > )
> > ```
> {: .solution}
{: .challenge}

## Variable creation functions
As mentioned before, any function that can take a vector of inputs and return a vector with the same 
length as an output can be used in a `mutate()` call. There are many R functions that can be used
in this situation, and `dplyr` introduces a number of new functions that may be useful for creating 
new variables. Some functions that you might find useful are:

#### Offset functions
To calculate differences between observations, you may be wanting to look at the value immediately
before or after it. These can be accessed using `lag()` or `lead()` respectively.
```{r}
mutate(gapminder, life_exp_prev = lag(lifeExp), life_exp_next = lead(lifeExp))
```

#### Cumulative computations
Cumulative sums (`cumsum()`), products (`cumprod()`) and means (`cummean()`) are all available to 
provide running computation of a variable. While not particularly useful for the gapminder data,
you can get an idea of how they function:

```{r mutate-cumsum}
mutate(gapminder, cumulative_life_exp = cumsum(lifeExp))
```

> ## Challenge 4
> Using an offset function, create a column showing the **change** in life expectancy between each sample period. Do any 
> countries have no change in life expectancy for a period?
> > ## Solution to Challenge 4
> > ``` {r mutate-sol-4}
> > diff_lifeExp <- mutate(gapminder, life_exp_change = lifeExp - lag(lifeExp))  
> >
> > filter(diff_lifeExp, life_exp_change == 0)
> > ```
> {: .solution}
{: .challenge}
