---
title: Cleaning Data 
teaching: 15
exercises: 30
questions:
  - How can I process data that is untidy/inconsistent/missing parts?
objectives:
  - Identify inconsistencies in data
  - Appropriately fix errors in data without overwriting the original
  - Reshape and manipulate data to make it tidy
  - Set data to missing where necessary
keypoints:
  - Real world data is messy
  - It takes time and care to prepare it for analysis and visualisation
source: Rmd
---

```{r, include=FALSE}
source("../bin/chunk-options.R")
knitr_fig_path("16-")

library(tidyverse)
```

Real world data sets rarely come in a form ready for analysis. Even the untidy example in the 
previous lesson was unusual because it was a complete data set with no missing values and informative
column names. Using `gather()`/`spread()` and `separate()`/`unite()` to reshape your data will solve 
a lot of problems, but they can not fix all the problems you will encounter. 

A few other situations you may encounter when trying to clean your data up for analysis are:

## Handling of missing data

Sometimes, missing data is not recorded as blank or`NA`, but as some other value. As discussed in the 
[data import]({{ page.root }}{% link _episodes/07-Reading-Data-In.md %}) lesson, you can 
define the missing data values on import with the `na` argument. But in some situations it might not
be possible because you are importing from a very specialised data format or using a data frame 
provided by someone else.

In this situation, we can recode values to be `NA` with the `na_if()`. You provide `na_if()` with 
a vector of values to check, and then the value that needs to be recoded as `NA`. This works 
well within a `mutate()` call.

```{r na-if, warning=F}
# -1 should be NA
missing <- data_frame(id = c("a", "b", "c"), value = c(10, -1, 20))

missing

missing %>% 
  mutate(value = na_if(value, -1))
```

## Filling in implied values

When recording a number of measurements, repeated values may be left blank for convenience, with the
implication that the first value is carried through the blank spaces. For example:

```{r implied-id}
implied_id <- data_frame(
  id = c("a", NA, NA, "b", NA, NA), 
  value = c(12, 23, 18, 34, 23, 16)
)

implied_id
```
Here, the `id` variable is recorded in the first row, but then left blank until an observation with 
a new id is recorded. To complete the data frame, you will need to `fill()` the `id` column, which 
replaces blank entries with the last recorded value.

```{r fill}
implied_id %>% 
  fill(id)
```

## Dealing with outliers/errors

Your data might have values that look like they have been recorded incorrectly or are outliers. It
can be appropriate to remove or alter these values in some situations. But just editing your raw data
is a poor choice because it does not leave any record of what you have done.

A conservative choice in this situation is to create a new variable containing your modifications, which
makes explicit what your criteria are. The `case_when()` function may be useful
for this. `case_when()` is a fast and powerful way of specifying an outcome in response to a set
of particular conditions. For example, assume we have the following data frame, but know that the 
measurement system can only record values between 0 and 30. We might decide that any measurements
outside these values should be set to either 0 or 30 (Is this a good way of dealing with these errors?).

```{r case-when}
outliers <- data_frame(
  id = c("a", "b", "c", "d", "e"),
  value = c(15, -2, 19, 9, 35)
)

outliers

# Clamp values outside of the range 0-30 to the closest limit
outliers %>% 
  mutate(clean_value = case_when(
    value < 0 ~ 0,
    value > 30 ~ 30,
    TRUE ~ value
  )
)
```

Each part of the `case_when()` call has two elements separated by a `~`. The left hand side is a 
*condition* that evaluates to `TRUE` or `FALSE`. The right hand side is the *output* that `case_when`
should return when the condition is `TRUE`. These conditions are evaluated from top to bottom and 
only return the output from the first condition that is `TRUE`.

So the code above can be read as:
 
 * Take the `outliers` data frame, then
 * `mutate` it to create a new column called `clean_value`
 * If the `value` column is less than zero, `clean_value` should be zero
 * Otherwise, if the `value` column is greater than 30, `clean_value` should be 30
 * Otherwise, if `TRUE` is `TRUE` (ie. always), `clean_value` should be the same as `value`



